---
title: "Emulation"
author: "JS"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE, fig.height = 3)
```

Disclaimer (5/8/26): this is still WIP, and some functionality may slightly change.

This example demonstrates some of the emulation functionality using a set of simulations from a pulmonary circulation model. Other aspects of this dataset are shown in `full_workflow.html`.

Load in design, data:

```{r, echo = TRUE}
setwd('..')
tmp <- getwd()
source(paste0(tmp, '/plot.R'))
source(paste0(tmp, '/basis.R'))
source(paste0(tmp, '/emulate.R'))
source(paste0(tmp, '/calibrate.R'))
load(paste0(tmp, '/data/SECRET_data.RData'))
# Currently need to load ggplot2 explicitly
library(ggplot2)
```


Scaling the design to [0,1] using the provided parameter ranges:

```{r}
scaled_design <- ScaleInputs(design[,1:4], parRanges, c(0,1))
```

We'll create 2 objects to demonstrate emulation. Firstly, calculate the SVD and coefficients:

```{r}
flow1_basis <- MakeDataBasis(flow1_data)
flow1_basis$grid <- data.frame(Time = seq(from=0,by = 0.01, len = 512))
flow1_coeffs <- GetEmData(scaled_design, flow1_basis, q = 5)
head(flow1_coeffs)
```

The dataset we've created here contains the 4 inputs, followed by the 5 leading basis coefficients, for each of the 100 simulations in turn.

And alternatively, select a subset of timepoints:

```{r}
time_inds <- c(50,150,300,500)
flow1_time <- data.frame(scaled_design, 
                         t50 = flow1_data[time_inds[1],],
                         t150 = flow1_data[time_inds[2],],
                         t300 = flow1_data[time_inds[3],],
                         t500 = flow1_data[time_inds[4],])
head(flow1_time)
```

Similarly, each row here is a different simulation, containing the 4 inputs, and the 4 quantities (timepoints) we've decided to emulate.

Splitting into train/validation for later:

```{r}
tt <- TrainTestSplit(scaled_design, seed = 58219)
train_inds <- tt$train_inds
val_inds <- tt$val_inds
```



## Default emulation

`FitEmulators` is the main function for training emulators.

If we just apply it to our data above, providing no other information and hence using the default settings:

```{r, results='hide'}
em_time <- FitEmulators(flow1_time[train_inds,])
em_coeffs <- FitEmulators(flow1_coeffs[train_inds,])
```

Comparing what's been fitted:

```{r}
summary(em_time)
em_time$response
```

```{r}
summary(em_coeffs)
unlist(lapply(1:5, function(k) em_coeffs[[k]]$response))
```

This highlights a key difference in the default behaviour:

  * If the provided data includes a column named `C1`, `FitEmulators` will automatically set all preceding columns as the emulator inputs, and will train emulators for `C1` **and all subsequent columns**. Here, `flow1_coeffs` contains the leading 5 coefficients, hence 5 emulators have been trained.
  
  * There's no column named `C1` in `flow1_time`. Given no other information, `FitEmulators` assumes the final column is the target output, and all preceding columns are inputs. Therefore, in the emulator `em_time` fitted above, `t50`, `t150` and `t300` have been included as predictors (along with the 4 actual inputs) for emulating `t500`:
  
```{r}
em_time$em
```

To instead get what we obviously intend here (emulate each of the 4 time points using the 4 input columns), we need to provide some extra information. First, we can point towards which columns are inputs:

```{r, results='hide'}
em_time <- FitEmulators(flow1_time[train_inds,], input = 1:4)
```

```{r}
summary(em_time)
em_time$response
```

This still fits a single emulator (for `t500` only), but now only uses the first 4 columns as inputs. 

To finally emulate all 4 outputs:

```{r, results='hide'}
em_time <- FitEmulators(flow1_time[train_inds,], input = 1:4, output = c('t50', 't150', 't300', 't500'))
```

```{r}
summary(em_time)
unlist(lapply(1:4, function(k) em_time[[k]]$response))
```

Alternatively, can select the outputs with column indices:

```{r, results='hide'}
em_time <- FitEmulators(flow1_time[train_inds,], input = 1:4, output_cols = 5:8)
```

```{r}
summary(em_time)
unlist(lapply(1:4, function(k) em_time[[k]]$response))
```



## Plotting emulators

The two main things we might want to immediately consider are leave-one-outs and predictions on some out-of-sample set.

We can produce these for an individual emulator:

```{r}
Validate(em_time[[1]], flow1_time[val_inds,])
```

```{r}
LeaveOneOut(em_time[[1]])
```

Or for multiple (recommended to store as a named object, as it will return each plot in a list, but also immediately print all in a single plot):

```{r}
val_time <- Validate(em_time, flow1_time[val_inds,])
summary(val_time)
```

```{r}
loo_time <- LeaveOneOut(em_time)
```

There's other plotting options, e.g., ordering by index instead:

```{r, fig.height=4}
val_time <- Validate(em_time, flow1_time[val_inds,], by_index = TRUE)
```

And/or by input:

```{r, results='hide'}
val_time <- Validate(em_time, flow1_time[val_inds,], by_index = TRUE, by_input = TRUE)
```

This doesn't automatically print all plots, but the `by_input` plots can be found via:

```{r, fig.height=4}
val_time[[1]]$input
```



## Refitting a specific emulator

By default, the same assumptions are used (currently) for each emulator fitted by a single call of `FitEmulator`. The same assumptions may not be be suitable across all outputs. Above, the out-of-sample predictions for `t150` aren't ideal (outside 95\% prediction intervals for 6 of the 25 points), so we may want to refit this with a different method.

We can replace the existing emulator for `t150` with a version refitted with different assumptions via:

```{r}
em_time <- RefitEmulator(em_time, 't150', method = 'het')
```

Instead of rgasp, we're tried fitting using hetGP instead. Reproducing the validation plots from before, we see the 2nd has been switched with the new emulator:

```{r, fig.height=4}
val_time <- Validate(em_time, flow1_time[val_inds,], by_index = TRUE)
```

At least when considering the 25 validation points, this has improved predictions slightly.



## Fitting different emulator types

We could directly compare different methods for emulating the same output. We could use `FitEmulators` as follows to fit for 4 different methods:

(Not run: `dgpsi`, used by `method = gp` and `method = dgp`, doesn't work currently in RMd. Running the below code in `R` as normal will work.)

```{r}
# emC1_v1 <- FitEmulators(flow1_coeffs[train_inds,], method = 'rgasp', input = 1:4, output = 'C1')
# emC1_v2 <- FitEmulators(flow1_coeffs[train_inds,], method = 'het', input = 1:4, output = 'C1')
# # emC1_v3 <- FitEmulators(flow1_coeffs[train_inds,], method = 'gp', input = 1:4, output = 'C1')
# emC1_v4 <- FitEmulators(flow1_coeffs[train_inds,], method = 'dgp', input = 1:4, output = 'C1')
```

And can then combine these into a single list to enable us to use the plotting functions as before:

```{r}
# emC1 <- list(emC1_v1, emC1_v2, emC1_v3, emC1_v4)
# val_C1 <- Validate(emC1, flow1_coeffs[val_inds,], by_index = TRUE)
```



## Different input sets

We could instead use different sets of inputs (more useful if emulating different outputs where there are different active sets of inputs, but as an illustration just done for `C1`):

```{r, results='hide'}
emC1_v1 <- FitEmulators(flow1_coeffs[train_inds,], method = 'rgasp', input = 4, output = 'C1')
emC1_v2 <- FitEmulators(flow1_coeffs[train_inds,], method = 'rgasp', input = 3:4, output = 'C1')
emC1_v3 <- FitEmulators(flow1_coeffs[train_inds,], method = 'rgasp', input = 2:4, output = 'C1')
emC1_v4 <- FitEmulators(flow1_coeffs[train_inds,], method = 'rgasp', input = 1:4, output = 'C1')
```


```{r}
emC1 <- list(emC1_v1, emC1_v2, emC1_v3, emC1_v4)
val_C1 <- Validate(emC1, flow1_coeffs[val_inds,], by_index = TRUE)
```

Training an emulator using only the 4th input is not particularly useful, and using the 3rd and 4th only doesn't do much better.


## Different covariance functions

A few covariance functions are currently possible (Matern 3/2, Matern 5/2, squared exponential). The default is currently Matern 5/2. Each underlying package names the covariance functions differently (below, written as `rgasp` accepts them), however `FitEmulators` accepts a range of naming conventions and converts them appropriately (see source code / help file for exhaustive list):

```{r, results='hide'}
emC1_v1 <- FitEmulators(flow1_coeffs[train_inds,], method = 'rgasp', input = 1:4, covariance = 'matern5_2', output = 'C1')
emC1_v2 <- FitEmulators(flow1_coeffs[train_inds,], method = 'rgasp', input = 1:4, covariance = 'matern3_2', output = 'C1')
emC1_v3 <- FitEmulators(flow1_coeffs[train_inds,], method = 'rgasp', input = 1:4, covariance = 'sexp', output = 'C1')
```

```{r}
emC1 <- list(emC1_v1, emC1_v2, emC1_v3)
val_C1 <- Validate(emC1, flow1_coeffs[val_inds,], by_index = TRUE)
```

## With/without nugget

Could also compare fitting the emulator with and without a nugget term. The default is to include one:

```{r, results='hide'}
emC1_v1 <- FitEmulators(flow1_coeffs[train_inds,], method = 'rgasp', input = 1:4, output = 'C1', nugget = TRUE)
emC1_v2 <- FitEmulators(flow1_coeffs[train_inds,], method = 'rgasp', input = 1:4, output = 'C1', nugget = FALSE)
```


```{r}
emC1 <- list(emC1_v1, emC1_v2)
val_C1 <- Validate(emC1, flow1_coeffs[val_inds,], by_index = TRUE)
```








