---
title: "Emulation"
author: "JS"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE, fig.height = 3)
```

Disclaimer (5/8/26): this is still WIP, and some functionality may slightly change.

This example demonstrates some of the emulation functionality using a set of simulations from a pulmonary circulation model. Other aspects of this dataset are shown in `full_workflow.html`.

Load in design, data:

```{r, echo = TRUE}
setwd('..')
tmp <- getwd()
source(paste0(tmp, '/plot.R'))
source(paste0(tmp, '/basis.R'))
source(paste0(tmp, '/emulate.R'))
source(paste0(tmp, '/calibrate.R'))
load(paste0(tmp, '/data/SECRET_data.RData'))
# Currently need to load ggplot2 explicitly
library(ggplot2)
```


Scaling the design to [0,1] using the provided parameter ranges:

```{r}
scaled_design <- ScaleInputs(design[,1:4], parRanges, c(0,1))
```

We'll create 2 objects to demonstrate emulation. Firstly, calculate the SVD and coefficients:

```{r}
flow1_basis <- MakeDataBasis(flow1_data)
flow1_basis$grid <- data.frame(Time = seq(from=0,by = 0.01, len = 512))
flow1_coeffs <- GetEmData(scaled_design, flow1_basis, q = 5)
head(flow1_coeffs)
```

The dataset we've created here contains the 4 inputs, followed by the 5 leading basis coefficients, for each of the 100 simulations in turn.

And alternatively, select a subset of timepoints:

```{r}
time_inds <- c(50,150,300,500)
flow1_time <- data.frame(scaled_design, 
                         t50 = flow1_data[time_inds[1],],
                         t150 = flow1_data[time_inds[2],],
                         t300 = flow1_data[time_inds[3],],
                         t500 = flow1_data[time_inds[4],])
head(flow1_time)
```

Similarly, each row here is a different simulation, containing the 4 inputs, and the 4 quantities (timepoints) we've decided to emulate.

Splitting into train/validation for later:

```{r}
tt <- TrainTestSplit(scaled_design, seed = 58219)
train_inds <- tt$train_inds
val_inds <- tt$val_inds
```



## Default emulation

`FitEmulators` is the main function for training emulators.

If we just apply it to our data above, providing no other information and hence using the default settings:

```{r, results='hide'}
em_time <- FitEmulators(flow1_time[train_inds,])
em_coeffs <- FitEmulators(flow1_coeffs[train_inds,])
```

Comparing what's been fitted:

```{r}
summary(em_time)
em_time$response
```

```{r}
summary(em_coeffs)
unlist(lapply(1:5, function(k) em_coeffs[[k]]$response))
```

This highlights a key difference in the default behaviour:

  * If the provided data includes a column named `C1`, `FitEmulators` will automatically set all preceding columns as the emulator inputs, and will train emulators for `C1` **and all subsequent columns**. Here, `flow1_coeffs` contains the leading 5 coefficients, hence 5 emulators have been trained.
  
  * There's no column named `C1` in `flow1_time`. Given no other information, `FitEmulators` assumes the final column is the target output, and all preceding columns are inputs. Therefore, in the emulator `em_time` fitted above, `t50`, `t150` and `t300` have been included as predictors (along with the 4 actual inputs) for emulating `t500`:
  
```{r}
em_time$em
```

To instead get what we obviously intend here (emulate each of the 4 time points using the 4 input columns), we need to provide some extra information. First, we can point towards which columns are inputs:

```{r, results='hide'}
em_time <- FitEmulators(flow1_time[train_inds,], input = 1:4)
```

```{r}
summary(em_time)
em_time$response
```

This still fits a single emulator (for `t500` only), but now only uses the first 4 columns as inputs. 

To finally emulate all 4 outputs:

```{r, results='hide'}
em_time <- FitEmulators(flow1_time[train_inds,], input = 1:4, output = c('t50', 't150', 't300', 't500'))
```

```{r}
summary(em_time)
unlist(lapply(1:4, function(k) em_time[[k]]$response))
```

Alternatively, we can select the outputs with column indices, or the inputs with column names:

```{r, results='hide'}
em_time <- FitEmulators(flow1_time[train_inds,], input = c('kMV','alpha','lrrA','lrrV'), output = 5:8)
```

```{r}
summary(em_time)
unlist(lapply(1:4, function(k) em_time[[k]]$response))
```

Any combination of `input`, `output` as numeric or character vectors are accepted, so long as any named columns and column indices exist in `tData`. If we try to select a column that doesn't exist:

```{r}
try(FitEmulators(flow1_time[train_inds,], input = 1:4, output = 'something'))
```

Selecting the same column as both an input and output is also not acceptable:

```{r}
try(FitEmulators(flow1_time[train_inds,], input = 1:4, output = 4))
```


Specifying `input` and `output` using column names rather than indices is generally more robust, and will flag if there's a missing name. Selecting using indices will only flag an issue if one or more is out-of-bounds, or if any are present in both the input and output set, but relies on the ordering of the training data being as expected.




## Plotting emulators

The two main things we might want to immediately consider are leave-one-outs and predictions on some out-of-sample set.

We can produce these for an individual emulator:

```{r}
Validate(em_time[[1]], flow1_time[val_inds,])
```

```{r}
LeaveOneOut(em_time[[1]])
```

Or for multiple (recommended to store as a named object, as it will return each plot in a list, but also immediately print all in a single plot):

```{r}
val_time <- Validate(em_time, flow1_time[val_inds,])
summary(val_time)
```

```{r}
loo_time <- LeaveOneOut(em_time)
```

There's other plotting options, e.g., ordering by index instead:

```{r, fig.height=4}
val_time <- Validate(em_time, flow1_time[val_inds,], by_index = TRUE)
```

And/or by input:

```{r, results='hide'}
val_time <- Validate(em_time, flow1_time[val_inds,], by_index = TRUE, by_input = TRUE)
```

This doesn't automatically print all plots, but the `by_input` plots can be found via:

```{r, fig.height=4}
val_time[[1]]$input
```



## Refitting a specific emulator

By default, the same assumptions are used for each emulator fitted by a single call of `FitEmulator`, which may not be desirable across all outputs we provide. Above, the out-of-sample predictions for `t150` aren't ideal (outside 95\% prediction intervals for 6 of the 25 points), so we may want to refit this with a different method.

We can replace the existing emulator for `t150` with a version refitted with different assumptions via:

```{r}
em_time <- RefitEmulator(em_time, 't150', method = 'het')
```

Instead of rgasp, we're tried fitting using hetGP instead. Reproducing the validation plots from before, we see the 2nd has been switched with the new emulator:

```{r, fig.height=4}
val_time <- Validate(em_time, flow1_time[val_inds,], by_index = TRUE)
```

At least when considering the 25 validation points, this has improved predictions slightly.



## Fitting different emulator types

We could directly compare different methods for emulating the same output. We could use `FitEmulators` as follows to fit for 4 different methods:

(Not run: `dgpsi`, used by `method = gp` and `method = dgp`, doesn't work currently in RMd. Running the below code in `R` as normal will work.)

```{r}
# emC1_v1 <- FitEmulators(flow1_coeffs[train_inds,], method = 'rgasp', input = 1:4, output = 'C1')
# emC1_v2 <- FitEmulators(flow1_coeffs[train_inds,], method = 'het', input = 1:4, output = 'C1')
# # emC1_v3 <- FitEmulators(flow1_coeffs[train_inds,], method = 'gp', input = 1:4, output = 'C1')
# emC1_v4 <- FitEmulators(flow1_coeffs[train_inds,], method = 'dgp', input = 1:4, output = 'C1')
```

And can then combine these into a single list to enable us to use the plotting functions as before:

```{r}
# emC1 <- list(emC1_v1, emC1_v2, emC1_v3, emC1_v4)
# val_C1 <- Validate(emC1, flow1_coeffs[val_inds,], by_index = TRUE)
```

It is not currently possible to fit emulators using multiple different packages in a single call of `FitEmulators`, and the above multiple calls and post-processing are required if we wish to do so. However, given a certain fitting method, we can specify different mean and covariance functions, nuggets, and active input sets for each output (see below).

Generally, it makes sense to use the same fitting method for all outputs initially, before refining any that need it later, via `RefitEmulator`. Prediction, validation etc. all work in the usual way, regardless of whether each individual emulator has been fitted with a different method.





## NEW - providing different assumptions

Now it's possible to provide different assumptions about the covariance function, nugget, input set, and mean function.

The functionality is that if we provide a vector of $k$ outputs to be emulated, we need to provide either:

  * A single choice of mean, covariance and nugget assumptions, and a single set of inputs, which are then used to fit each emulator
  
  * A vector of length $k$ of mean and/or covariance and/or nugget assumptions, and a list of $k$ input sets.
  
All combinations of the above are possible. We could choose a specific covariance function for each of the $k$ emulators, but set `nugget = TRUE` to universally assume this, and use the same set of inputs.


### Different input sets

To use different sets of input parameters across outputs, we can provide a list that contains either a character or numeric vector relating to the columns of the training dataset, relating to each chosen output.

This may be useful because we're emulating different outputs where there are different active sets of inputs, or perhaps we wish to compare different assumptions for emulating the same output. I'll do the latter here, and fit a constant mean emulator to `C1` with different sets of inputs included.

If we provide a list of input sets, but just a single output, this gives an error as 4 input sets have been provided, but only 1 output (if this were the other way around, with 4 outputs and a single input set, as above this will assume each output has the same set of inputs, and will work fine. The general behaviour is to duplicate input assumptions where needed, but not to automatically duplicate outputs):

```{r}
inputs <- colnames(scaled_design)
try(FitEmulators(flow1_coeffs[train_inds,], method = 'rgasp', input = list(inputs[4],inputs[3:4],inputs[2:4],inputs), output = 'C1'))
```

Instead, explicitly repeating the same output 4 times means that each input set (and any other combination of assumptions we provide) will be used to emulate `C1` each time:

```{r}
emC1 <- FitEmulators(flow1_coeffs[train_inds,], method = 'rgasp', input = list(inputs[4],inputs[3:4],inputs[2:4],inputs), output = rep('C1',4))
```

```{r}
val_C1 <- Validate(emC1, flow1_coeffs[val_inds,], by_index = TRUE)
```

Training an emulator using only the 4th input is not particularly useful, and using the 3rd and 4th only doesn't do much better.


### Different covariance functions

A few covariance functions are currently possible (Matern 3/2, Matern 5/2, squared exponential), with the default currently Matern 5/2. Each underlying package names the covariance functions differently (below, written as `rgasp` accepts them), however `FitEmulators` accepts a range of naming conventions and converts them appropriately (see source code / help file for exhaustive list).

Here emulating `C1` with each of the 3 covariance functions, keeping all other assumptions the same:

```{r}
emC1 <- FitEmulators(flow1_coeffs[train_inds,], method = 'rgasp', input = inputs, covariance = c('matern5_2', 'matern3_2', 'sexp'), output = rep('C1', 3))
val_C1 <- Validate(emC1, flow1_coeffs[val_inds,], by_index = TRUE)
```

### With/without nugget

Could also compare fitting the emulator with and without a nugget term. The default is to include one:

```{r}
emC1 <- FitEmulators(flow1_coeffs[train_inds,], method = 'rgasp', input = inputs, nugget = c(TRUE,FALSE), output = rep('C1',2))
val_C1 <- Validate(emC1, flow1_coeffs[val_inds,], by_index = TRUE)
```



## NEW - editing mean functions

Mean functions can be provided in a similar way. The possible options are:

  * `constant`: assumes a constant mean;
  
  * `linear`: assumes a mean that is linear in the inputs;
  
  * `formula`: any other formula can be given, provided each of the inputs included in it are named in the training data frame. Any formula that can be provided to `lm` can be used here. The response and `~` should not be included.

Training an emulator to `C1` with 4 different mean function arguments. The 2nd and 3rd are identical as an illustration:

```{r}
mean_fn_vector <- c('constant',
                  'linear',
                  'lrrV+kMV+alpha+lrrA',
                  'I(kMV^2) + I(alpha^2)')

emC1 <- FitEmulators(flow1_coeffs[train_inds,], method = 'rgasp', input = inputs, output = rep('C1',4), mean_fn = mean_fn_vector)
val_C1 <- Validate(emC1, flow1_coeffs[val_inds,], by_index = TRUE)
loo_C1 <- LeaveOneOut(emC1, by_index = TRUE)
```

Should get identical predictions for the 2nd and 3rd:

```{r}
preds <- Predict(emC1, flow1_coeffs[val_inds,])
head(preds$Expectation)
head(preds$Variance)
```

Should also get identical predictions if re-order columns of test data:

```{r}
preds2 <- Predict(emC1, flow1_coeffs[val_inds,9:1])
summary(c(preds$Expectation - preds2$Expectation))
summary(c(preds$Variance - preds2$Variance))
```