---
title: "SECRET example"
author: "JS"
output:
  html_document:
    toc: yes
  pdf_document:
    toc: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE, cache = TRUE, fig.height = 3)
```

Disclaimer (5/8/26): this is still WIP, and some functionality may slightly change.

A worked example using the pulmonary model from the SECRET competition.

Load in design, data:

```{r, echo = TRUE}
setwd('..')
tmp <- getwd()
source(paste0(tmp, '/plot.R'))
source(paste0(tmp, '/basis.R'))
source(paste0(tmp, '/emulate.R'))
source(paste0(tmp, '/calibrate.R'))
load(paste0(tmp, '/data/SECRET_data.RData'))
# Currently need to load ggplot2 explicitly
library(ggplot2)
```

Parameter ranges, and hence scaling the design to [0,1]:

```{r}
parRanges
scaled_design <- ScaleInputs(design[,1:4], parRanges, c(0,1))
```


## Single 1D profile

First let's consider only the first output, and process the data into a `DataBasis` object:

```{r}
DataBasis <- MakeDataBasis(flow1_data)
```

We can now plot various aspects using the functionality of the package. Plotting the full ensemble:

```{r}
PlotData(DataBasis)
```

Plotting the leading 4 basis vectors:

```{r}
PlotBasis(DataBasis, q = 4)
```

Everything will work fine and treat inputs as 1:512, but by assigning a data frame to `$grid`, we can relabel:

```{r}
DataBasis$grid <- data.frame(Time = seq(from=0, by = 0.01, len = 512))
PlotData(DataBasis)
```


## Multiple 1D profiles

```{r}
DataBasis_multi <- MakeDataBasis(pulm_data)
PlotData(DataBasis_multi)
```

This is not a particularly useful visualisation, but by assigning time/variable names in a grid, the same code will automatically plot according to this:

```{r}
DataBasis_multi$grid <- data.frame(Time = 1:512,
                              Output = c(rep('Flow1', 512), rep('Flow2', 512), rep('Pressure', 512)))
PlotData(DataBasis_multi)
```

Or the basis:

```{r}
PlotBasis(DataBasis_multi, q = 4)
```



## Emulation

As an initial demonstration, select a few particular timepoints from Flow1. Combining the (scaled) inputs with the outputs for 4 different timepoints:

```{r}
flow1_time <- data.frame(scaled_design, 
                         t50 = flow1_data[50,],
                         t150 = flow1_data[150,],
                         t300 = flow1_data[300,],
                         t500 = flow1_data[500,])
```

Splitting into training and validation sets:

```{r}
tt <- TrainTestSplit(scaled_design, seed = 58219)
train_inds <- tt$train_inds
val_inds <- tt$val_inds
```

More details are provided in `emulate.html`, but briefly, we can fit a single emulator by providing a single named output, and pointing to which columns we want to be treated as inputs:

```{r, results='hide'}
em_t150 <- FitEmulators(flow1_time[train_inds,], input = 1:4, output = 't150')
```

```{r}
summary(em_t150)
```

This has stored an `rgasp` emulator under `$em`, what mean function was used (here none, hence `NULL`), and also what the training data was (including only the selected inputs and output, rather than the full given `tData`).

To look at the emulator predictions, we could plot leave one outs:

```{r}
LeaveOneOut(em_t150, by_index = TRUE)
```

Or predict over the validation set:

```{r}
Validate(em_t150, flow1_time[val_inds,], by_index = TRUE)
```

We could also plot vs the inputs:

```{r}
Validate(em_t150, flow1_time[val_inds,], by_input = TRUE)
```


## Emulate multiple outputs

The same code can be used to emulate all 4 of the chosen outputs independently:

```{r}
em_time <- FitEmulators(flow1_time[train_inds,], input = 1:4, output = c('t50', 't150', 't300', 't500'))
```

Now the fitted `em_time` is a list of 4 individual emulators:

```{r}
summary(em_time)
summary(em_time[[1]])
```

The `Validate` and `LeaveOneOut` functions work with lists of emulators as well:

```{r}
loo_time <- LeaveOneOut(em_time, by_index = TRUE)
```

```{r}
val_time <- Validate(em_time, flow1_time[val_inds,], by_index = TRUE)
```



## Predictions

Predictions work the same for each type of emulator. Providing a set of 5000 points sampled from input space:

```{r}
set.seed(58219)
BigDesign <- lhs::randomLHS(5000,4)
colnames(BigDesign) <- colnames(scaled_design)
preds_t150 <- Predict(em_t150, BigDesign)
preds_time <- Predict(em_time, BigDesign)
```

The `Predict` function outputs `$Expectation` and `$Variance`, with the number of rows equal to the number of prediction points (here, 5000), and the number of columns equal to the number of emulators: 

```{r}
summary(preds_t150)
summary(preds_time)

dim(preds_t150$Expectation)
dim(preds_time$Expectation)
```



## Calibration

Let's assume some 'observations'. Run 100 from the initial design wasn't used in training the emulators, so we'll set this as the target:

```{r}
obs <- flow1_data[,100]
truth <- scaled_design[100,]
```

We can calculate the implausibility for an individual output, given a set of predictions, and given observation error variance (here, set at 0.5) and discrepancy variance (defaults to zero, as could provide combined with the observation error):

```{r}
impl_t150 <- Impl(preds_t150, Obs = obs[150], ObsVar = 0.5, DiscVar = 0, kmax = 1)
```

Or for multiple outputs (independently):

```{r}
impl_time <- Impl(preds_time, Obs = obs[c(50,150,300,500)], ObsVar = 0.5, DiscVar = 0, kmax = 1)
```

Where we have now provided 4 sets of predictions and 4 observations. `ObsVar` (and `DiscVar`) has still been set as a single value, which is interpreted as each observation has the same observational error. Instead, this could be provided as a vector of length 4:

```{r}
#impl_time <- Impl(preds_time, Obs = obs[c(50,150,300,500)], ObsVar = c(0.5, 0.1, 1, 0.5), DiscVar = c(0, 1, 0.2, 0), kmax = 1)
```

Although prediction and calculating implausibility can be done by hand as above, usually it will be more convenient to combine them:

```{r}
hm_time <- HistoryMatch(em_time, BigDesign, Obs = obs[c(50,150,300,500)], ObsVar = 0.5, kmax = 1)
summary(hm_time)
```

There's several outputs returned by this function:

  * `$Design` is the same as the given design, but with a column appended indicating whether each point is in NROY space or not.
  
  * `$Preds` contains the emulator predictions at the given design.
  
  * `$Impl`, `$NROY`, `$Pass_kth`, `$Pass_N`, `$inNROY` and `$wave` are related to the calibration part. The implausibility itself, for each provided input vector and each observation:

```{r}
head(hm_time$Impl)
```

The proportion of space not ruled out:

```{r}
hm_time$NROY
```

The proportion of space that satisfies the kth constraint:

```{r}
hm_time$Pass_kth
```

The proportion of space that passes 0, 1, 2, 3, 4 out of 4 constraints:

```{r}
hm_time$Pass_N
```

Whether each row of the design is in NROY:

```{r}
head(hm_time$inNROY)
```

As a demonstration that what's happening within the `HistoryMatch` function is identical to what we did previously, the predictions are the same:

```{r}
summary(hm_time$Preds$Expectation - preds_time$Expectation)
```

As is the implausibility:

```{r}
summary(hm_time$Impl - impl_time$Impl)
```

Taking the output, we can visualise the NROY space in 1D and 2D, with the truth overlaid:

```{r, fig.height = 5}
PlotNROY(hm_time$Design, k = 1:4, Truth = truth)
```

A multiwave example, using both scalar outputs and higher dimensional, correlated outputs, is found in `systemic.html`.





## Emulate full output

Now we'll use dimension reduction to emulate the full Flow1 output curve.

We already have the required `DataBasis` object (calculated at the start to simplify plotting):

```{r}
summary(DataBasis)
```

This already contains the projections of the (centered) data onto the basis vectors:

```{r}
dim(DataBasis$Coeffs)
```

As an illustration, we'll select the leading 5 basis vectors arbitrarily, but more or fewer might perform better. Combining these coefficients with the design:

```{r}
tDataCoeffs <- GetEmData(scaled_design, DataBasis, q = 5)
head(tDataCoeffs)
```

Training emulators for the leading 5 coefficients works in the same way as above. In fact, by default, if the code finds a column named `C1`, it will automatically treat this as the first output to emulate, and also train emulators to all subsequent columns. Hence, the below is equivalent to setting `emulate_names = c('C1', 'C2', 'C3', 'C4', 'C5')`:

```{r}
emCoeffs <- FitEmulators(tDataCoeffs[train_inds,], input = 1:4)
```

We have 5 emulators, one for each of the given sets of coefficients:

```{r}
summary(emCoeffs)
unlist(lapply(1:5, function(k) emCoeffs[[k]]$response))
```

Looking at LOOs and out-of-sample predictions:

```{r}
valCoeffs <- Validate(emCoeffs, tDataCoeffs[val_inds,], by_index = TRUE)
```

```{r}
looCoeffs <- LeaveOneOut(emCoeffs, by_index = TRUE)
```

We seem to be still predicting quite well for the 5th coefficient, which suggests at least trying to emulate the next, rather than treating as noise.

Prediction is as above:

```{r}
predsCoeffs <- Predict(emCoeffs, BigDesign)
head(predsCoeffs$Expectation)
```


## Calibration for the full field

As for the set of 4 individual metrics above, the function `HistoryMatch` will handle the high-dimensional calibration. Let's do it by hand first to be clearer about what's happening. The only difference is that instead of the 1D implausibility `Impl`, we require `ImplField`:

```{r}
implFull <- ImplField(predsCoeffs, DataBasis, Obs = obs, ObsVar = 0.5)
summary(implFull$Impl)
implFull$threshold
implFull$NROY
summary(implFull$inNROY)
```

This function exploits the basis structure of the emulator, and in fact calculate the full 512D implausibility as the sum of the reconstruction error (fixed) and the 5D coefficient implausibility:

```{r}
ReconError(obs - DataBasis$EnsembleMean, DataBasis$tBasis[,1:5], scale = FALSE)
```

(Actually, this isn't *quite* right, because in fact calculate it after accounting for basis uncertainty. But same general idea.)

As a check that the above decomposition is correct (it is), there's a function that calculates the full implausibility by mapping the emulator expectation and variance back to the field. Here, this is tractable as we only have a $512\times512$ inversion for each input, but still quite slow, so only do for a few:

```{r}
preds_tmp <- Predict(emCoeffs, BigDesign[1:100,])
implFull_v2 <- ImplFieldRecon(preds_tmp, DataBasis, obs, ObsVar = 0.5)
summary(implFull$Impl[1:100] - implFull_v2)
```

Instead directly using the `HistoryMatch` function:

```{r}
hmCoeffs <- HistoryMatch(emCoeffs, BigDesign, DataBasis, Obs = obs, ObsVar = 0.5)
summary(hmCoeffs$Impl)
hmCoeffs$NROY
```

The results are exactly the same as when running `Predict` and `ImplField` separately. Plotting:

```{r, fig.height = 5}
PlotNROY(hmCoeffs$Design, k = 1:4, Truth = truth)
```

Although the observations are the same, these results aren't fully comparable to those when using 4 metrics, as now using all 512 outputs, and those we didn't consider before may provide different information.

For more on working with the dimension reduction and coefficient emulator code, see XXXX.









```{r, eval=FALSE, echo=FALSE}
By default, this function subtracts the ensemble mean, and then calculates the SVD across the (centred) data. `$Q` and `$Lambda` are only stored if this function is used for weighted SVD, hence are not returned here (and indeed, not required in future calculations).

Plotting the leading few vectors:
```


```{r, eval=FALSE, echo=FALSE}
# q <- 9
# plot_basis <- data.frame(Time = rep(1:t, q),
#                          Vector = rep(1:q, each = t),
#                          Weight = c(DataBasis$tBasis[,1:q]))
# ggplot(plot_basis, aes(Time, Weight)) +
#   geom_line() +
#   facet_wrap(vars(Vector))


Generally we truncate this basis for some small $q$ (such that a large enough amount of variability has been explained and/or such that any later vectors are just noise, with no signal from the parameters).


# q <- ExplainT(DataBasis, vtot = 0.95)
# q


Here, 4 vectors explain 95\% of the variability in the data. In practice, we might want to try emulating the 5th, 6th etc. to see if these are predictable, but let's just use the 1st 4 for now.

Alternatively, plot cumulative variance explained:


# vars <- lapply(1:n, function(k) VarExplained(DataBasis$tBasis[,1:k], DataBasis$CentredField))

# ggplot(data.frame(q = 1:n, Proportion = unlist(vars)), aes(x = q, y = Proportion)) +
#   geom_line() +
#   ylim(0,1)




Projecting the data onto the basis:


# Coeffs <- Project(data = DataBasis$CentredField, 
#                   basis = DataBasis$tBasis[,1:q])
# colnames(Coeffs)[1:q] <- paste("C",1:q,sep="")
# summary(Coeffs)




# tDataC <- data.frame(tData[,1:4], # getting the scaled version from before
#                      Noise = runif(n, -1, 1), 
#                      Coeffs)
# head(tDataC)


Now we can build emulators for these $q$ coefficients. We can do this simultaneously for all $q$ (as long as we're happy using the same assumptions for each). To make things consistent across each emulator, I'm going to define training/validation sets by hand, and then fit to the full training set by setting `training_prop = 1` (rather than allowing the function to do this split for me):


# set.seed(321)
# inds <- sample(1:n, n)
# train_inds <- inds[1:75]
# val_inds <- inds[-c(1:75)]
# train_data <- tDataC[train_inds,]
# val_data <- tDataC[val_inds,]




# em_coeffs <- BasisEmulators(tDataC, q, mean_fn = 'step', maxdf = 5, training_prop = 1)

# em_coeffs <- BasisEmulators(train_data, q, mean_fn = 'step', maxdf = 5, training_prop = 1)


The output here is now a list of $q$ emulator objects, e.g.


# summary(em_coeffs[[1]])
# summary(em_coeffs[[2]])


Now in `ValidateGasp`, we need to provide the validation set (it's no longer internal to the `BasisEmulator` object):


# par(mfrow=c(2,2), mar=c(4,2,2,2))
# ValidateGasp(em_coeffs[[1]], val_data)
# ValidateGasp(em_coeffs[[2]], val_data)
# ValidateGasp(em_coeffs[[3]], val_data)
# ValidateGasp(em_coeffs[[4]], val_data)
# 
# par(mfrow=c(2,2), mar=c(4,2,2,2))
# LeaveOneOut(em_coeffs[[1]]);LeaveOneOut(em_coeffs[[2]]);LeaveOneOut(em_coeffs[[3]]);LeaveOneOut(em_coeffs[[4]])



## Prediction

The validation plots above are performing prediction across the test data. In general, we can predict for any sets of inputs, for either a 1D emulator, or for a set of basis emulators. Doing so across a space-filling design in parameter space (here, we are still working in [-1,1]):


# ns <- 1000 # usually want more, but set low for speed
# BigDesign <- 2*as.data.frame(randomLHS(ns, 4)) - 1 
# colnames(BigDesign) <- colnames(design)[1:4]
# 
# Preds_1D <- PredictGasp(BigDesign, em3)
# Preds_basis <- BasisPredGasp(BigDesign, em_coeffs)


These store slightly different things - in the 1D case, we get mean/sd/lower95/upper95 (the same as what `predict.rgasp` returns):


# summary(Preds_1D)


In the basis case, I only store `$Expectation` and `$Variance` (as this is all we need for history matching, and because prediction intervals can be derived from these, so don't want to store these if we have a lot of emulators):


# summary(Preds_basis)


Both of these objects are $ns \times q$ dimensional, where each row corresponds to an input we're predicting at, and each column corresponds to a basis vector:


# dim(Preds_basis$Expectation)
# dim(Preds_basis$Variance)



## Reconstructing

From basis coefficients (whether given by projection, or predicted by an emulator), we can reconstruct a prediction of the original field.

Let's consider the runs from the validation set. First produce predictions for them:


# Preds_val <- BasisPredGasp(val_data, em_coeffs)


Let's just take the first run, and compare the mean/variance to the truth (in coefficient/projection space):


# data.frame(Truth = as.numeric(val_data[1,-c(1:5)]),
#            Mean = Preds_val$Expectation[1,],
#            Var = Preds_val$Variance[1,])


Plot the true run vs emulator reconstruction of it (still with the ensemble mean removed, as this will dominate):


# plot_data <- data.frame(Time = 1:t,
#                         Truth = DataBasis$CentredField[,val_inds[1]],
#                         Recon = Recon(Preds_val$Expectation[1,], DataBasis$tBasis[,1:q]))
# plot_data2 <- melt(plot_data, id.vars = c('Time'))
# ggplot(plot_data2, aes(Time, value, col = variable)) + geom_line()


Looks quite accurate (remember this run was not used for fitting the emulator).

Mean prediction by itself is not that informative, also sample from the emulators:


# em_samp <- matrix(0, t, 100)
# for (s in 1:100){
#   samp <- rnorm(q, 
#                 mean = Preds_val$Expectation[1,],
#                 sd = sqrt(Preds_val$Variance[1,]))
#   rec <- Recon(samp, DataBasis$tBasis[,1:q])
#   em_samp[,s] <- rec
# }

# ggplot(plot_data2, aes(Time, value, col = variable)) + 
#   geom_line(data = data.frame(Time = 1:t, value = c(em_samp), s = rep(1:100, each = t)), aes(Time, value, linetype = as.factor(s)), col = 'grey', alpha = 0.6) +
#   geom_line(size = 1.25) +
#   scale_linetype_manual(values = rep(1,100), guide = 'none')


Really, we should also add the uncertainty due to the discarded basis vectors - by truncating the basis after $q$ vectors, we know that we're not perfectly reconstructing the true simulations via the emulators, and need to acknowledge this additional variability. It is likely small relative to the variability we've explained (less than 5\%) but we should still account for it.

There's a function that will calculate this variance matrix, and we can sample from it:


# extra_var <- DiscardedBasisVariance(DataBasis, q)
# extra_var_samples <- rmvnorm(100, rep(0, t), extra_var)
# dim(extra_var_samples)
# 
# plot_samples <- data.frame(Time = 1:512,
#                            epsilon = c(t(extra_var_samples)),
#                            s = rep(1:100, each = t))
# 
# ggplot(plot_samples, aes(Time, epsilon, col = as.factor(s))) + 
#   geom_line(alpha = 0.6) +
#   theme(legend.position = 'none')


The above is 100 samples from the discarded vectors, and we have clear correlated structure in these (some of the later basis vectors are likely quite noisy, however these will have lower variance/eigenvalues, so are drowned out by the earlier, more structured, discarded vectors).

Adding this variability onto our emulator samples:


# ggplot(plot_data2, aes(Time, value, col = variable)) + 
#   geom_line(data = data.frame(Time = 1:t, value = c(em_samp + t(extra_var_samples)), s = rep(1:100, each = t)), aes(Time, value, linetype = as.factor(s)), col = 'grey', alpha = 0.6) +
#   geom_line(size = 1.25) +
#   scale_linetype_manual(values = rep(1,100), guide = 'none')


Before, we were missing the data in places, because the truncated basis did not have the ability to perfectly reconstruct the truth. Now, the truth lies within our uncertainty.

Just plotting 95\% prediction intervals for clarity:


# plot_data <- data.frame(Time = 1:t,
#                         Truth = DataBasis$CentredField[,val_inds[1]],
#                         Recon = rep(Recon(Preds_val$Expectation[1,], DataBasis$tBasis[,1:q]), 2),
#                         Lower = c(apply(em_samp, 1, quantile, probs = 0.025), apply(em_samp + t(extra_var_samples), 1, quantile, probs = 0.025)),
#                         Upper = c(apply(em_samp, 1, quantile, probs = 0.975), apply(em_samp + t(extra_var_samples), 1, quantile, probs = 0.975)),
#                         Type = rep(c('EmVar', 'FullVar'), each = t))
# plot_data2 <- melt(plot_data, id.vars = c('Time', 'Type'))
# pal <- scales::hue_pal()(2)
# ggplot(plot_data2, aes(Time, value, col = variable, linetype = variable)) +
#   geom_line(size = 0.8) +
#   facet_wrap(vars(Type)) +
#   scale_linetype_manual(values = c(1,1,2,2)) +
#   scale_colour_manual(values = pal[c(1,2,2,2)]) +
#   theme(legend.position = 'none')





Or doing for 4 different inputs (immediately adding in the variance from the discarded vectors):


# em_samp <- array(0, dim = c(t, 100, 4))
# plot_data <- NULL
# for (i in 2:5){
#   plot_data <- rbind(plot_data,
#                      data.frame(Time = 1:t,
#                                 Truth = DataBasis$CentredField[,val_inds[i]],
#                                 Recon = Recon(Preds_val$Expectation[i,], DataBasis$tBasis[,1:q]),
#                                 Run = i))
#   for (s in 1:100){
#     samp <- rnorm(q, 
#                   mean = Preds_val$Expectation[i,],
#                   sd = sqrt(Preds_val$Variance[i,]))
#     rec <- Recon(samp, DataBasis$tBasis[,1:q])
#     em_samp[,s,i-1] <- rec
#   }
# }
# 
# plot_data2 <- melt(plot_data, id.vars = c('Time', 'Run'))
# 
# extra_var_samples <- t(rmvnorm(400, rep(0, t), extra_var))
# dim(extra_var_samples) <- c(t, 100, 4)
# 
# plot_data_samp <- data.frame(Time = 1:t, 
#                              value = c(em_samp + extra_var_samples), 
#                              s = rep(1:100, each = t),
#                              Run = rep(2:5, each = t*100))
# 
# ggplot(plot_data2, aes(Time, value, col = variable)) +
#   facet_wrap(vars(Run)) +
#   geom_line(data = plot_data_samp, aes(Time, value, linetype = as.factor(s)), col = 'grey', alpha = 0.6) +
#   geom_line(size = 0.8) +
#   scale_linetype_manual(values = rep(1,100), guide = 'none')



## Visualising response surface

We may want to see what's happening across the output space is we systematically change values of the inputs. Here, the 2 dominant parameters are `kMV` and `alpha`, so let's vary these.

Before we predicted over a large Latin hypercube. Now let's systematically vary these 2 parameters, and fix the others at the centre of their range (zero):


# ns <- 1000
# pred_seq <- seq(from = -1, to = 1, by = 0.04)
# pred_grid <- expand.grid(kMV = pred_seq, alpha = pred_seq)
# NewDesign <- data.frame(pred_grid, lrrA = 0, lrrV = 0)
# PredsNewBasis <- BasisPredGasp(NewDesign, em_coeffs)


Plotting the expectation of `C1`:


# ggplot(data.frame(NewDesign, C1 = PredsNewBasis$Expectation[,1]),
#        aes(x = kMV, y = alpha, col = C1)) +
#   geom_point(size = 3, shape = 15) +
#   scale_colour_viridis()


Or `C2`:


# ggplot(data.frame(NewDesign, C2 = PredsNewBasis$Expectation[,2]),
#        aes(x = kMV, y = alpha, col = C2)) +
#   geom_point(size = 3, shape = 15) +
#   scale_colour_viridis()


This is a 2D surface in the 4D input space with `lrrA` and `lrrV` both fixed to zero. Instead setting them both to 1 demonstrates different behaviour in the output:


# ns <- 1000
# pred_seq <- seq(from = -1, to = 1, by = 0.04)
# pred_grid <- expand.grid(kMV = pred_seq, alpha = pred_seq)
# NewDesign <- data.frame(pred_grid, lrrA = 1, lrrV = 1)
# PredsNewBasis <- BasisPredGasp(NewDesign, em_coeffs)
# ggplot(data.frame(NewDesign, C1 = PredsNewBasis$Expectation[,1]),
#        aes(x = kMV, y = alpha, z = C1, col = C1)) +
#   geom_point(size = 3, shape = 15) +
#   geom_contour(breaks = c(-5), col = 'white') +
#   scale_colour_viridis(limits = c(-85,85)) 



Depending on what we're interested in (relationship on this 2D surface for specific values of other parameters, vs general behaviour integrated across all other uncertain inputs), instead  of fixing the other inputs we could sample. For example, for each specific pair of $(kMV, alpha)$ values (each point on the plot), we could sample a Latin hypercube across the remaining parameters, average the emulator prediction across these points, and plot this instead.
```
